Combining network-guided GWAS to discover susceptibility mechanisms for breast cancer

### 2020-03-26 manuscript/manuscript.tex
comments on discussion

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the \caz{results}{outputs/solutions} of multiple network methods \caz{on the gene-gene interaction network}{}. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \footnotesize{
    \item \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  }
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\subsection{The consensus overcomes the problems of the individual solutions}
\label{results:drawbacks}

@@

By virtue of using a SNP subnetwork, SConES analyzes each SNP in their context. It therefore selects SNPs in genes none of whose interactors are associated to the disease, as well as SNPs in non-coding regions or in non-interacting genes. In fact, due to linkage disequilibrium, such genes are favored by SConES, as selecting SNPs in an LD block which overlaps with a gene favors selecting the rest of the gene. This might explain why SConES produces similar results on the GS and GM networks, heavily affected by linkage disequilibrium (Supplementary Figure~\ref{sfig:pearson_methods}). On the other hand, SConES penalizes selecting SNPs and not their neighbors. This makes it conservative regarding SNPs with many interactions, for instance those mapped to hubs in the PPIN. For this reason, SConES GI did not select any protein coding gene, despite selecting similar regions as SConES GS (Figure~\ref{fig:issues}C). In fact SConES GS and SConES GM select regions related to breast cancer, like 16q12 (\emph{TOX3}, Section~\ref{results:conventional}), 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), and 10q26 (\emph{FGFR2}, Section~\ref{results:conventional}). On top of that only SConES GS selects region 8q24 (\emph{POU5F1B} \cite{breyer_expressed_2014}). We hypothesize that the lack of results on the PPIN network of SConES GI and LEAN are due to the same cause: the absence of joint association of a module. Although in the case of SConES other hyperparameters could lead to a more informative solution (e.g. lower \(\lambda\), Section~\ref{methods:methods}), it is unclear what is the best strategy to find them. In addition, due to the iCOGS SNP array design, the genome of GENESIS participants has not been unbiasedly surveyed: some regions are fine-mapped --- which might distort gene structure in GM and GI networks --- while others are under studied --- hurting the accuracy with which the GS network captures the genome structure.

@@

We compared the results of different network methods to the European cohort of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date in breast cancer (Section~\ref{methods:bcac}). This comparison is pertinent as, despite not targeting the same population(s), we expect a shared genetic architecture at the gene level % \cite{required}
TODO CAN'T FIND A CITATION FOR THIS, at which most examined network metods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly (Fisher's exact test P-value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) (Figure~\ref{fig:solution_overview}C). However, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

In this article we evaluate the viability of a systems biology take on genetic studies by examining a GWAS dataset on familial breast cancer. Such an approach addresses two of the largest issues with GWAS: interpretability and an overly conservative statistical framework that hinders discovery. This is achieved by considering the biological context of each of the genes and SNPs. Based on divergent considerations of what the desired set of biomarkers is, several methods for network-guided biomarker discovery have been proposed. We reviewed six of them. Most of them produced a relevant subset of biomarkers, although the solutions were notably different. In one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, which risk oversimplifying their richness. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Despite these differences, all the solutions produced comparably poor discrimination capabilities on a linear classifier trained on them, and were remarkably unstable, and hence likely to produce change in face of slightly different inputs.

To overcome the problems posed by the individual methods while exploiting their strengths, we propose combining them into a consensus subnetwork. We use a straightforward aggregation to generate it, including any node that was recovered by at least two methods. The resulting network is a synthesis of the altered mechanism: it is smaller than the largest solutions (HotNet2, SigMod and dmGWAS), which makes it more manageable, and includes the majority of the strongly associated smaller solutions (SConES and heinz). The consensus subnetwork captures mechanisms and genes known to be related to cancer, recovering known breast cancer susceptibility genes as well as genome regions associated to breast cancer susceptibility. However, thanks to its smaller size and its network structure, it provides compelling hypotheses of non-canonical mechanisms involved in carcinogenesis, like mitochondrial translation and chaperone activity.

The strength of network-based analyses comes from leveraging prior knowledge to boost discovery. In consequence, they show their shortcomings in front of understudied genes, especially those not in the network. Out of the 32\,767 genes that we can map the genotyped SNPs to, 60.7\% (19\,887) are not in the protein-protein interaction network. The majority of those (14\,660) are non-coding genes, mainly lncRNA, miRNA, and snRNA (Supplementary Figure~\ref{sfig:biotypes_excluded}). Yet, RNA genes like \emph{CASC16} are associated to breast cancer (Section~\ref{results:conventional}), reminding us of the importance of using networks beyond coding genes. Even protein-coding genes linked to breast cancer susceptibility \cite{ahmed_newly_2009}, like \emph{NEK10} (P-value 1.6 \texttimes{} 10\textsuperscript{-5}, located near \emph{SLC4A7}) or \emph{POU5F1B}, were absent from the network. However, on average protein-coding genes absent from the PPIN are less associated with this phenotype (Wilcoxon rank-sum P-value~=~2.79 \texttimes{} 10\textsuperscript{-8}, median P-values of 0.43 and 0.47). As we are using interactions from high-throughput experiments, such difference cannot be due to well-known genes having more known interactions. As disease genes tend to be more central \cite{pinero_uncovering_2016}, we hypothesize that it is due to interactions between central genes being more likely. It is worth noting that network approaches that do not use PPIs, like SConES GS and GM, did recover SNPs in \emph{NEK10} and \emph{CASC16}. This shows the potential of SNP networks, in which SNPs are linked when there is evidence of co-function, to perform network-guided GWAS even in the absence of gene-level interactions. Lastly, all the methods rely heavily on how SNPs are mapped to genes. In Section~\ref{results:consensus} we highlight ambiguities that appear when genes overlap or are in linkage disequilibrium.

As not all databases compile the same interactions, the choice of the PPIN determines the final output. In this work we used exclusively interactions from HINT from high-throughput experiments. This responds to concerns of some authors about adding interactions identified in targeted studies and prone to a ``rich getting richer'' phenomenon: popular genes have a higher proportion of their interactions described \cite{cai_broker_2010,das_hint:_2012}. Their presense might bias the results of the presented methods, as they change the topology of the network by adding nodes closer to other nodes than average. On the other hand, \citet{huang_systematic_2018} found that the best predictor of the performance of a network for disease gene discovery is the size of the network, which supports using the largest amount of interactions. When we compared the impact of using a larger network containing interactions from both high-throughput experiment and the literature (Section~\ref{methods:networks}), we found that for most of the methods it did not greatly change the size or the stability of the solution, the classification accuracy, or the runtime (Supplementary Figure~\ref{sfig:lc_ht_comparison}). This supports using only interactions from high-throughput experiments, which produces apparently similar solutions and avoids falling into ``circular reasonings'', where the best known genes are artificially pushed into the selected solutions.

A crucial step for the gene based methods is the computation of the gene score. In this work we used VEGAS2v2 \cite{mishra_vegas2:_2015} due to the flexibility it offers to use user-specified gene annotations. However, it presents known problems (selection of an appropriate percentage of top SNPs, long runtimes and P-value precision limited to the number of permutations \cite{nakka_gene_2016}), and other algorithms might have more statistical power. Another important decision is how to handle LD in a GWAS. VEGAS2v2 accounts for LD patterns, and hence an LD pruning step would not impact gene-based network methods, although it would speed up VEGAS2v2's computation time. With regards to SConES, less SNPs would lead to simpler SNP networks and, possibly, shorter runtimes. However, as mentioned in Section~\ref{results:drawbacks}, LD patterns seem paramount to SConES' solutions, and an LD pruning step could potentially alter them.

>>>>

\subsection{Network methods successfully identify genes associated with breast cancer}
\label{results:separate_networks}
\begin{table}[htbp]
  \centering
  \begin{threeparttable}
\caption{\label{tab:gene_solutions}
Summary statistics on the \caz{results}{outputs|solutions} of multiple network methods \caz{on the gene-gene interaction network}{}. The first row contains the summary statistics on the whole network.}
\centering
\begin{tabular}{lrrrrr}
Network & \# genes & \# edges & \mean{Betweenness} & \median{P}\textsubscript{gene} & \(\rho\)\textsubscript{consensus}\\
\hline
HINT HT & 13\,619 & 142\,541 & 16\,706 & 0.46 & 0.066\\
\hline
Consensus & 55 & 117 & 74\,062 & 0.0051 & 1\\
dmGWAS & 194 & 450 & 49\,115 & 0.19 & 0.41\\
heinz & 4 & 3 & 113\,633 & 0.0012 & 0.21\\
HotNet2 & 440 & 374 & 7\,739 & 0.048 & 0.31\\
LEAN & 0 & 0 & - & - & -\\
SConES GI & 0 (1) & 0 & - & - & -\\
SigMod & 142 & 249 & 92\,603 & 0.0083 & 0.73\\
\end{tabular}
\begin{tablenotes}
  \footnotesize{
    \item \textbf{\emph{\# genes}}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median P-value of the selected genes. \textbf{\(\rho\)\textsubscript{consensus}}: Pearson correlation between the subnetwork and the consensus network.
  }
\end{tablenotes}
\end{threeparttable}
\end{table}

@@

\subsection{The consensus overcomes the limitations of the individual solutions}
\label{results:drawbacks}
\cazcom{}{I replaced ``problems'' with ``limitations'' in the section title.}

@@

By virtue of using a SNP subnetwork, SConES analyzes each SNP in their context. It therefore selects SNPs in genes none of whose interactors are associated to the disease, as well as SNPs in non-coding regions or in non-interacting genes. In fact, due to linkage disequilibrium, such genes are favored by SConES, as selecting SNPs in an LD block which overlaps with a gene favors selecting the rest of the gene. This might explain why SConES produces similar results on the GS and GM networks, heavily affected by linkage disequilibrium (Supplementary Figure~\ref{sfig:pearson_methods}). On the other hand, SConES penalizes selecting SNPs and not their neighbors. This makes it conservative regarding SNPs with many interactions, for instance those mapped to hubs in the PPIN. For this reason, SConES GI did not select any protein coding gene, despite selecting similar regions as SConES GS (Figure~\ref{fig:issues}C). In fact SConES GS and SConES GM select regions related to breast cancer, like 16q12 (\emph{TOX3}, Section~\ref{results:conventional}), 3p24 (\emph{SLC4A7}/\emph{NEK10} \cite{ahmed_newly_2009}), 5p12 (\emph{FGF10}, \emph{MRPS30} \cite{quigley_5p12_2014}), and 10q26 (\emph{FGFR2}, Section~\ref{results:conventional}). On top of that only SConES GS selects region 8q24 (\emph{POU5F1B} \cite{breyer_expressed_2014}). We hypothesize that the lack of results on the PPIN network of SConES GI and LEAN are due to the same cause: the absence of joint association of a module. Although in the case of SConES other hyperparameters could lead to a more informative solution (e.g. lower \(\lambda\), Section~\ref{methods:methods}), it is unclear what is the best strategy to find them. In addition, due to the iCOGS SNP array design, the genome of GENESIS participants has not been unbiasedly surveyed: some regions are fine-mapped --- which might distort gene structure in GM and GI networks --- while others are under studied --- \caz{hurting}{limiting|hindering} the accuracy with which the GS network captures the genome structure.

@@

We compared the results of different network methods to the European cohort of the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017}, the largest GWAS to date in breast cancer (Section~\ref{methods:bcac}). This comparison is pertinent as, \caz{despite not targeting}{although they do not target} the same population(s), we expect a shared genetic architecture at the gene level % \cite{required}
\cazcom{TODO CAN'T FIND A CITATION FOR THIS}{Maybe we don't need to justify it? Or maybe Nadine/Fabienne can provide something?}, at which most examined network metods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provides a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC, in an equivalent way to the one described in Section~\ref{methods:node_score}. The solutions provided by the different network approaches overlap significantly \caz{}{with BCAC findings} (Fisher's exact test P-value \textless~0.019). The gene-based methods achieve comparable precision (2\%-25\%) and recall (1.3-12.1\%) (Figure~\ref{fig:solution_overview}C). However, while SConES GI at the SNP-level achieves a similar recall (8.6\%), it shows a much higher precision (47.3\%).

@@

In this article we evaluate the viability of a systems biology take on genetic \caz{}{association} studies by examining a GWAS dataset on familial breast cancer. Such an approach addresses two of the largest issues with GWAS: interpretability and an overly conservative statistical framework that hinders discovery. This is achieved by \caz{considering}{using interaction networks to consider} the biological context of each of the genes and SNPs. \cazcom{}{I'd like to say here that this is in line with the recently proposed omnigenic model, itself supported by a large body of work on using networks to understand disease. I would also like to say that although some folks use the omnigenic model to claim that GWAS don't work, such network methods are a way of bridging classical GWAS and the omnigenic model.}  Based on \caz{divergent}{various|diverse} considerations of what the desired set of biomarkers is, several methods for network-guided biomarker discovery have been proposed \caz{}{in the literature|in recent years}. We reviewed six of them.

Most of these methods produced a relevant subset of biomarkers, although the solutions were notably different. \caz{In}{At} one end of the spectrum, SConES and heinz preferred small, highly associated solutions, providing a conveniently short list of biomarkers, at the expense of not shedding much light on the etiology of the disease. On the other end, SigMod and dmGWAS gravitate towards larger, less associated solutions which provide a wide overview of the biological context. While this deepens our understanding of the disease and provide biological hypotheses, they require further analyses, \cazcom{which risk oversimplifying their richness}{Are you arguing that these methods give richer insights but that we're not equipped to analyze them, resulting in poorer insights? I find this a bit convoluted...}. HotNet2 balances both approaches at the expense of producing the largest solution: a constellation of many, highly associated, small subnetworks. Despite these differences, all solutions \caz{produced comparably poor discrimination capabilities on a linear classifier trained on them}{have equally poor discriminative power|are equally poor at discriminating cases from controls}\caz{, and were remarkably unstable, and hence likely to produce change in face of slightly different inputs.}{. In addition, all methods are remarkably unstable, yielding different solutions for slightly different inputs.}
\cazcom{}{I think we need to discuss this. Why aren't we able to classify cases from controls (multiple non-genetic factors, disease heterogeneity...)? Why is our approach still useful if we aren't, in fact, discovering biomarkers? (it points at relevant pathways, etc) Why are the results unstable? (correlations in the data i.e. everything is connected i.e. the omnigenic model?) Why can we still do something with the results if they are unstable? (not clear to me at this stage, if we don't have stability at the pathway/subnetwork level or for the consensus network)}

To overcome the \caz{problems posed by}{limitations of} the individual methods while exploiting their strengths, we propose combining them into a consensus subnetwork. We use a straightforward aggregation \caz{to generate it}{}, including any node that was recovered by at least two methods. The resulting network is a synthesis of the altered mechanism\caz{}{s}: it is smaller than the largest solutions (HotNet2, SigMod and dmGWAS), which makes it more manageable, and includes the majority of the strongly associated smaller solutions (SConES and heinz). The consensus subnetwork captures mechanisms and genes known to be related to cancer, recovering known breast cancer susceptibility genes as well as genome regions associated to breast cancer susceptibility. However, thanks to its smaller size and its network structure, it provides compelling hypotheses of non-canonical mechanisms involved in carcinogenesis, like mitochondrial translation and chaperone activity.

The strength of network-based analyses comes from leveraging prior knowledge to boost discovery. In consequence, they show their shortcomings in front of understudied genes, especially those not in the network. Out of the 32\,767 genes that we can map the genotyped SNPs to, 60.7\% (19\,887) are not in the protein-protein interaction network. The majority of those (14\,660) are non-coding genes, mainly lncRNA, miRNA, and snRNA (Supplementary Figure~\ref{sfig:biotypes_excluded}). Yet, RNA genes like \emph{CASC16} are associated to breast cancer (Section~\ref{results:conventional}), reminding us of the importance of using networks beyond coding genes. In addition, even protein-coding genes linked to breast cancer susceptibility \cite{ahmed_newly_2009}, like \emph{NEK10} (P-value 1.6 \texttimes{} 10\textsuperscript{-5}, located near \emph{SLC4A7}) or \emph{POU5F1B}, were absent from the network. However, on average protein-coding genes absent from the PPIN are less associated with this phenotype (Wilcoxon rank-sum P-value~=~2.79 \texttimes{} 10\textsuperscript{-8}, median P-values of 0.43 and 0.47). As we are using interactions from high-throughput experiments, such difference cannot be due to well-known genes having more known interactions. As disease genes tend to be more central \cite{pinero_uncovering_2016}, we hypothesize that it is due to interactions between central genes being more likely. It is worth noting that network approaches that do not use PPIs, like SConES GS and GM, did recover SNPs in \emph{NEK10} and \emph{CASC16}. \cazcom{}{Say a word about their ability to recover SNPs that have not been mapped to genes, i.e. mostly intergenic genes?} This shows the potential of SNP networks, in which SNPs are linked when there is evidence of co-function, to perform network-guided GWAS even in the absence of gene-level interactions. Lastly, all the methods rely heavily on \cazcom{how SNPs are mapped to genes}{maybe discuss the promising possibility of creating more complex mapping, relying on eQTLs and contact maps?}. In Section~\ref{results:consensus} we highlight ambiguities that appear when genes overlap or are in linkage disequilibrium.

As not all databases compile the same interactions, the choice of the PPIN determines the final output. In this work we used exclusively interactions from HINT from high-throughput experiments. This responds to concerns \caz{of some authors}{} about adding interactions identified in targeted studies and prone to a ``rich getting richer'' phenomenon: popular genes have a higher proportion of their interactions described \cite{cai_broker_2010,das_hint:_2012}. Their presense might bias the results of the presented methods, as they change the topology of the network by adding nodes closer to other nodes than average. On the other hand, \citet{huang_systematic_2018} found that the best predictor of the performance of a network for disease gene discovery is the size of the network, which supports using the largest amount of interactions. When we compared the impact of using a larger network containing interactions from both high-throughput experiment and the literature (Section~\ref{methods:networks}), we found that for most of the methods it did not greatly change the size or the stability of the solution, the classification accuracy, or the runtime (Supplementary Figure~\ref{sfig:lc_ht_comparison}). This supports using only interactions from high-throughput experiments, which produces apparently similar solutions and avoids falling into ``circular reasonings'', where the best known genes are artificially pushed into the selected solutions.

A crucial step for the gene based methods is the computation of the gene score. In this work we used VEGAS2v2 \cite{mishra_vegas2:_2015} due to the flexibility it offers to use user-specified gene annotations. However, it presents known problems (selection of an appropriate percentage of top SNPs, long runtimes and P-value precision limited to the number of permutations \cite{nakka_gene_2016}), and \cazcom{other algorithms}{cite a few among SKAT, MAGMA, MAGENTA, PEGASUS etc.} might have more statistical power. Another important decision is how to handle LD in a GWAS. VEGAS2v2 accounts for LD patterns, and hence an LD pruning step would not impact gene-based network methods, although it would speed up VEGAS2v2's computation time. With regards to SConES, less SNPs would lead to simpler SNP networks and, possibly, shorter runtimes. However, as mentioned in Section~\ref{results:drawbacks}, LD patterns seem paramount to SConES' solutions, and an LD pruning step could potentially alter them.
